---
title: "DIIG Data Challenge"
author: "Alayna Binder"
date: "2023-09-14"
output: html_document
---

## Context

The World Health Organization, as part of a recent initiative to reduce early deaths in developing countries, wants to learn more about what factors have the strongest correlation with life expectancy. Specifically, they would like to confidently fund specific sectors within developing countries that will have the overall greatest impact on improvements in life expectancy. You have been tasked with examining this data to provide guidance for WHO on where you believe they should focus their efforts on.

## Guiding Questions

* Which variables have the strongest positive and negative correlation to high life expectancy?
* Are there any similarities in variable outcomes based on Developed vs Developing countries?
* Are there similar trends when considering various columns related to vaccinations vs income/wealth vs deaths?
* If a disease like Hepatitis B is present at high levels in certain countries, is there correlation with other diseases as well?

_NOTE: Guiding questions are meant to help you get started and brainstorm. You do not need to specifically answer these questions for the presentation. Feel free to explore anything that is of interest to you in the dataset._

```{r load-pkg-data, echo = FALSE}
#| message: FALSE
library(readr)
library(dplyr)
library(tidyverse)
library(tidymodels)
library(naniar) # Missing variable summary
library(car) # VIF
library(MASS) # Backward selection
library(glmnet) # LASSO regression
library(caret) # Split test/training
library(ggcorrplot)
lifeexp <- read_csv("Life Expectancy Data.csv")
```

## Data

```{r data-processing, echo = FALSE}

lifeexp_developing <- lifeexp |>
  filter(Status == 'Developing') |>
  mutate(Status = case_when(Status == 'Developing' ~ 0)) |>
  na.omit(Population)

# ----------------------------------------------------------------------------

lifeexp_developed <- lifeexp |>
  filter(Status == 'Developed') |>
  mutate(Status = case_when(Status == 'Developed' ~ 1)) |>
  na.omit(Population)

# ----------------------------------------------------------------------------

lifeexp <- lifeexp |>
  mutate(Status = case_when(Status == 'Developing' ~ 0,
                               Status == 'Developed' ~ 1)) |>
  na.omit(Population)

lifeexp$Status = as.factor(lifeexp$Status)
```

```{r correlation-multicollinearity, echo = FALSE}
# Correlation matrix to see aliased coefficients
lifeexp_cor <- lifeexp |>
  dplyr::select(-c(Country, Status))

lifeexp_cor = cor(lifeexp_cor)
# Correlated variables (>0.90): GDP/expenditure, under-five deaths/infant deaths, thinness 1-19/thinness 5-9, income composition/schooling
# Expenditure is used to calculate GDP

# ----------------------------------------------------------------------------

# Multicollinearity
lifeexp_mc <- lifeexp |>
  dplyr::select(-c(Country, `infant deaths`, GDP, `thinness  1-19 years`))

fit_lifeexp <- lm(`Life expectancy` ~ ., data = lifeexp_mc)
t(t(vif(fit_lifeexp)))
summary(fit_lifeexp)

# ----------------------------------------------------------------------------

# Remove correlated/multicollinear variables
lifeexp_lm <- lifeexp |>
  dplyr::select(-c(`infant deaths`, GDP, `thinness  1-19 years`, `Income composition of resources`))

lifeexp_lm_developing <- lifeexp_developing |>
  dplyr::select(-c(`infant deaths`, GDP, `thinness  1-19 years`, `Income composition of resources`))

lifeexp_lm_developed <- lifeexp_developed |>
  dplyr::select(-c(`infant deaths`, GDP, `thinness  1-19 years`, `Income composition of resources`))
```

``` {r data-exploration, echo = FALSE}
# Missing variable summary using naniar package
miss_var_summary(lifeexp)
miss_var_summary(lifeexp_developing)
miss_var_summary(lifeexp_developed)

# Glimpse
glimpse(lifeexp)
```

### Variables

* Status: Developed or Developing country.
* Year: There are several data points for many countries so consider various years!
* Adult Mortality: Probability of dying between 15 - 60 years per 1000 population.
* Infant Deaths: Number of infant deaths per 1000 population.
* Alcohol: Alcohol, recorded per capita (15+) consumption (in liters of pure alcohol).
* Percentage Expenditure: Expenditure on health as a % of GDP.
* Hepatitis B/Polio/Diphtheria: HepB/Polio/Diphtheria immunization coverage among 1-year-olds.
* Measles: Number of cases per 1000 population.
* BMI: Average BMI of the entire population.
* Total Expenditure: General government expenditure on health as a percentage of total government expenditure.
* HIV/AIDS: Deaths per 1000 live births HIV/AIDS (0-4 years)
* Thinness: Prevalence of thinness among children and adolescents for age interval.
* Income Composition: Human Development Index in terms of income composition of resources (index ranging from 0 to 1.)
* Schooling: Number of years of schooling.

`infant deaths`, `GDP`, and `thinness  1-19 years` were removed from the data set because they resulted in multicollinearity/were highly correlated with `under-five deaths` (0.997), `percentage expenditure` (0.960), and `thinness 5-9 years` (0.928), respectively. `Income composition of resources` was also removed due to a correlation greater than 0.70. This was chosen over `Schooling` because years of schooling is a more specific item the WHO can target.

### Missing Data

In practice, we want to avoid using variables that have more than 10% missingness to avoid bias. In this data set, `Population` (22.2%), `Hepatitis B` (18.8%), and `GDP` (15.2%) all have more than 10% missingness. Data imputation, however, is not appropriate in this scenario because the data is not missing at random. Certain countries have missing data in several variable columns, sometimes over a span of several years, so imputing data may cause misleading results. Since this analysis is focused on reviewing general factors and not specific countries, missing values are dropped. Dropping the NAs in population removes all NAs in the data set.

## Exploratory Data Analysis

```{r developed-vs-developing, echo = FALSE, results = 'hide'}
lifeexp_bar <- lifeexp |>
  mutate(Status = case_when(Status == 0 ~ 'Developing',
                            Status == 1 ~ 'Developed'))

ggplot(lifeexp_bar, aes(x = Status)) +
  geom_bar() +
  labs(x = "Status",
       y = "Count",
       title = "Distribution of Status")

# Before dropping NAs of population: 500 developed, 2400 developing
# After dropping: 250 developed, 1400 developing
```

```{r density-matrix, echo = FALSE, results = 'hide'}
lifeexp_density <- lifeexp |>
  dplyr::select(-c(Country, Status))

# Convert your dataframe to long format for plotting
lifeexp_long <- lifeexp_density %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value")

# Create a matrix of density plots
ggplot(lifeexp_long, aes(x = Value)) +
  geom_density(fill = "blue", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  theme_minimal() +
  labs(title = "Matrix of Density Plots")
```

```{r population-vs-lifeexp, echo = FALSE, warning = FALSE, results = 'hide'}
ggplot(lifeexp, aes(x = Population, y = `Life expectancy`)) +
  geom_point() +
  xlim(0, 260000) + # Remove outliers
    labs(y = "Life Expectancy", 
       title = "Population vs. Life Expectancy")

# There is no pattern of association between population and life expectancy
```

```{r disease-scatterplot, echo = FALSE, results = 'hide'}
data_to_plot <- data.frame(
  LifeExpectancy = lifeexp_lm_developing$`Life expectancy`,
  Measles = lifeexp_lm_developing$Measles,
  HIV_AIDS = lifeexp_lm_developing$`HIV/AIDS`,
  Diphtheria = lifeexp_lm_developing$Diphtheria,
  HepatitisB = lifeexp_lm_developing$`Hepatitis B`,
  Polio = lifeexp_lm_developing$Polio
)

# Create a scatterplot matrix with all variables
scatter_matrix <- ggplot(data_to_plot, aes(y = LifeExpectancy)) +
  geom_point(aes(x = Measles), color = "red") +
  geom_point(aes(x = HIV_AIDS), color = "blue") +
  geom_point(aes(x = Diphtheria), color = "green") +
  geom_point(aes(x = HepatitisB), color = "purple") +
  geom_point(aes(x = Polio), color = "orange") +
  labs(x = "Disease Variables", y = "Life Expectancy") +
  ggtitle("Scatterplot Matrix: Life Expectancy vs. Diseases")

# Print the scatterplot matrix
print(scatter_matrix)
```

```{r disease-corrplot, echo = FALSE}
lifeexp_cor_developing <- lifeexp_lm_developing |>
  dplyr::select(c(`Hepatitis B`, `HIV/AIDS`, Measles, Polio, Diphtheria))

lifeexp_cor_developing = cor(lifeexp_cor_developing)

ggcorrplot(lifeexp_cor_developing, hc.order = TRUE, type = 'lower')
```

Explains why initial disease scatter plot is not an informative visual.

```{r disease-scatterplot2, echo = FALSE}
lifeexp_disease_scatter <- lifeexp_lm |>
  mutate(Status = case_when(Status == 0 ~ 'Developing',
                            Status == 1 ~ 'Developed'))

sp <- ggplot(lifeexp_disease_scatter,
       aes(x = `HIV/AIDS`, y = `Life expectancy`, color = Status)) +
  geom_point() +
  labs(x = "HIV/AIDS Deaths (per 1000 Live Births)",
       y = "Life Expectancy",
       title = "HIV/AIDS vs. Life Expectancy") +
  guides(color = guide_legend(reverse=TRUE))

sp + scale_color_manual(values = c("steelblue2", "indianred2"))
```

```{r thinness-scatterplot, echo = FALSE}
lifeexp_thinness_scatter <- lifeexp_lm |>
  mutate(Status = case_when(Status == 0 ~ 'Developing',
                            Status == 1 ~ 'Developed'))

sp2 <- ggplot(lifeexp_thinness_scatter,
       aes(x = `thinness 5-9 years`, y = `Life expectancy`, color = Status)) +
  geom_point() +
  xlim(0, 23) + # Remove outliers
  labs(x = "Thinness 5-9 Years",
       y = "Life Expectancy",
       title = "Thinness 5-9 Years vs. Life Expectancy") +
  guides(color = guide_legend(reverse=TRUE)) +
  theme(legend.position = "bottom")

sp2 + scale_color_manual(values = c("steelblue2", "indianred2"))
```

```{r alcohol-scatterplot, echo = FALSE, results = 'hide'}
lifeexp_alcohol_scatter <- lifeexp_lm |>
  mutate(Status = case_when(Status == 0 ~ 'Developing',
                            Status == 1 ~ 'Developed'))

sp3 <- ggplot(lifeexp_alcohol_scatter,
       aes(x = Alcohol, y = `Life expectancy`, color = Status)) +
  geom_point() +
  labs(x = "Alcohol Consumption (in liters of pure alcohol)",
       y = "Life Expectancy",
       title = "Alcohol Consumption vs. Life Expectancy") +
  guides(color = guide_legend(reverse=TRUE)) +
  theme(legend.position = "bottom")

sp3 + scale_color_manual(values = c("steelblue2", "indianred2"))
```

```{r immunization-vs-death, echo = FALSE, results = 'hide'}
lifeexp_alcohol_scatter <- lifeexp_lm |>
  mutate(Status = case_when(Status == 0 ~ 'Developing',
                            Status == 1 ~ 'Developed'))

sp4 <- ggplot(lifeexp_alcohol_scatter,
       aes(x = `Hepatitis B`, y = `under-five deaths`, color = Status)) +
  geom_point() +
  labs(x = "Hepatitis B Immunization Coverage",
       y = "Under-five Deaths",
       title = "Hepatitis B Immunization Coverage vs. Under-five Deaths") +
  guides(color = guide_legend(reverse=TRUE)) 

sp4 + scale_color_manual(values = c("steelblue2", "indianred2"))
```

```{r schooling-viz, echo = FALSE, results = 'hide'}
lifeexp_school_scatter <- lifeexp_lm |>
  mutate(Status = case_when(Status == 0 ~ 'Developing',
                            Status == 1 ~ 'Developed'))

sp5 <- ggplot(lifeexp_school_scatter,
       aes(x = Schooling, y = `percentage expenditure`, color = Status)) +
  geom_point() +
  labs(x = "Years of Schooling",
       y = "Percentage Expenditure\non Health (as a % of GDP)",
       title = "Years of Schooling vs. Percentage Expenditure") +
  guides(color = guide_legend(reverse=TRUE))

sp5 + scale_color_manual(values = c("steelblue2", "indianred2"))

# ----------------------------------------------------------------------------

sp6 <- ggplot(lifeexp_school_scatter,
       aes(x = Schooling, fill = Status)) +
  geom_density(aes(color = Status), alpha = 0.5) +  # Use color aesthetic
  labs(x = "Years of Schooling",
       y = "Density",
       title = "Density Distribution of Years of Schooling") +
  guides(fill = guide_legend(reverse = TRUE)) + 
  scale_fill_manual(values = c("Developing" = "indianred2", "Developed" = "steelblue2")) +
  scale_color_manual(values = c("Developing" = "indianred2", "Developed" = "steelblue2")) +
  guides(color = FALSE)

sp6

# ----------------------------------------------------------------------------

sp7 <- ggplot(lifeexp_school_scatter,
       aes(x = Schooling, y = `Life expectancy`, color = Status)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "slateblue4") +
  labs(x = "Years of Schooling",
       y = "Life Expectancy",
       title = "Years of Schooling vs. Life Expectancy") +
  guides(color = guide_legend(reverse=TRUE))

sp7 + scale_color_manual(values = c("steelblue2", "indianred2"))
```

```{r thinness-viz}

sp8 <- ggplot(lifeexp_school_scatter,
       aes(x = `thinness 5-9 years`, fill = Status)) +
  geom_density(aes(color = Status), alpha = 0.5) +  # Use color aesthetic
  labs(x = "Prevalence of Thinness from 5-9 Years",
       y = "Density",
       title = "Density Distribution of Thinness Prevalence") +
  guides(fill = guide_legend(reverse = TRUE)) + 
  scale_fill_manual(values = c("Developing" = "indianred2", "Developed" = "steelblue2")) +
  scale_color_manual(values = c("Developing" = "indianred2", "Developed" = "steelblue2")) +
  guides(color = FALSE)

sp8

# ----------------------------------------------------------------------------

sp9 <- ggplot(lifeexp_school_scatter,
       aes(x = `thinness 5-9 years`, y = `Adult Mortality`, color = Status)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "slateblue4") +
  labs(x = "Prevalence of Thinness",
       y = "Adult Mortality",
       title = "Prevalence of Thinness vs. Adult Mortality") +
  guides(color = guide_legend(reverse=TRUE))

sp9 + scale_color_manual(values = c("steelblue2", "indianred2"))
```

```

## Model Selection

```{r forward-selection, echo = FALSE}
# Define starting model
start.mod <- lm(`Life expectancy` ~ 1 , data = lifeexp_lm)

# Define stopping model 
stop.mod <- formula(lm(`Life expectancy` ~ ., data = lifeexp_lm))

# Forward selection
alpha.crit <- 0.2
ffit.lifeexp <- stats::step(start.mod,
             scope=stop.mod,
             direction="forward",
             test="F",
             trace = 0,
             k=qchisq(1-alpha.crit,1))

summary(ffit.lifeexp)
# lm(formula = `Life expectancy` ~ Country + Year + `HIV/AIDS` + 
#     Schooling + `thinness 5-9 years` + Alcohol + `Hepatitis B` + 
#     `under-five deaths`, data = lifeexp_lm)

# ----------------------------------------------------------------------------

start.mod.developing <- lm(`Life expectancy` ~ 1 , data = lifeexp_lm_developing)

stop.mod.developing <- formula(lm(`Life expectancy` ~ ., data = lifeexp_lm_developing))

alpha.crit <- 0.2
ffit.lifeexp.developing <- stats::step(start.mod.developing,
             scope=stop.mod.developing,
             direction="forward",
             test="F",
             trace = 0,
             k=qchisq(1-alpha.crit,1))

summary(ffit.lifeexp.developing)
# lm(formula = `Life expectancy` ~ Country + Year + `HIV/AIDS` + 
#     Schooling + Alcohol + `thinness 5-9 years` + `under-five deaths` + 
#     `Adult Mortality`, data = lifeexp_lm_developing)

# ----------------------------------------------------------------------------

start.mod.developed <- lm(`Life expectancy` ~ 1 , data = lifeexp_lm_developed)

stop.mod.developed <- formula(lm(`Life expectancy` ~ ., data = lifeexp_lm_developed))

alpha.crit <- 0.2
ffit.lifeexp.developed <- stats::step(start.mod.developed,
             scope=stop.mod.developed,
             direction="forward",
             test="F",
             trace = 0,
             k=qchisq(1-alpha.crit,1))

summary(ffit.lifeexp.developed)
# lm(formula = `Life expectancy` ~ Country + Year + `thinness 5-9 years` + 
#     `Adult Mortality`, data = lifeexp_lm_developed)
```

```{r backward-selection, echo = FALSE}
# Define model
back.mod <- lm(`Life expectancy` ~ ., data = lifeexp_lm)

# Backward selection
bfit.lifeexp <- back.mod |>
  stepAIC(direction='backward', trace=FALSE)

summary(bfit.lifeexp)
# lm(formula = `Life expectancy` ~ Country + Year + Alcohol + `Hepatitis
#     B` + `under-five deaths` + `HIV/AIDS` + `thinness 5-9 years` + Schooling, data = lifeexp_lm)

# ----------------------------------------------------------------------------

back.mod.developing <- lm(`Life expectancy` ~ ., data = lifeexp_lm_developing)

bfit.lifeexp.developing <- back.mod.developing |>
  stepAIC(direction='backward', trace=FALSE)

summary(bfit.lifeexp.developing)
# lm(formula = `Life expectancy` ~ Country + Year + `Adult Mortality` + 
#     Alcohol + `under-five deaths` + `HIV/AIDS` + `thinness 5-9 years` + Schooling, data = lifeexp_lm_developing)

# ----------------------------------------------------------------------------

back.mod.developed <- lm(`Life expectancy` ~ ., data = lifeexp_lm_developed)

bfit.lifeexp.developed <- back.mod.developed |>
  stepAIC(direction='backward', trace=FALSE)

summary(bfit.lifeexp.developed)
# lm(formula = `Life expectancy` ~ Country + Year + `thinness 5-9 years`, 
#     data = lifeexp_lm_developed)
```

```{r stepwise-selection, echo = FALSE}
# Fit the full model 
stepwise.mod <- lm(`Life expectancy` ~., data = lifeexp_lm)

# Stepwise regression
step.model <- stepAIC(stepwise.mod, direction = "both", 
                      trace = FALSE)

summary(step.model)
# lm(formula = `Life expectancy` ~ Country + Year + Alcohol + `Hepatitis B` + 
#     `under-five deaths` + `HIV/AIDS` + `thinness 5-9 years` + Schooling, data = lifeexp_lm)

# ----------------------------------------------------------------------------

stepwise.mod.developing <- lm(`Life expectancy` ~., data = lifeexp_lm_developing)

step.model.developing <- stepAIC(stepwise.mod.developing, direction = "both", 
                      trace = FALSE)

summary(step.model.developing)
# lm(formula = `Life expectancy` ~ Country + Year + `Adult Mortality` + 
#     Alcohol + `under-five deaths` + `HIV/AIDS` + `thinness 5-9 years` + Schooling, data = lifeexp_lm_developing)

# ----------------------------------------------------------------------------

stepwise.mod.developed <- lm(`Life expectancy` ~., data = lifeexp_lm_developed)

step.model.developed <- stepAIC(stepwise.mod.developed, direction = "both", 
                      trace = FALSE)

summary(step.model.developed)
# lm(formula = `Life expectancy` ~ Country + Year + `thinness 5-9 years`, 
#     data = lifeexp_lm_developed)
```

General model: all three selection models (forward, backward, both) output the same variables.

Notable variation between developing and developed subsets, but all selected variables for the developing subset were the same.

```{r lasso-regression, echo = FALSE, results = 'hide'}
#| eval: false

lifeexp_lasso <- lifeexp |>
  dplyr::select(-c(Country, Status))

yy <- lifeexp_lasso[,'Life expectancy']
xx <- as.matrix(lifeexp_lasso[,c("Year", "Adult Mortality", "Alcohol", "percentage expenditure", "Hepatitis B", "Measles", "BMI", "under-five deaths", "Polio", "Total expenditure", "Diphtheria", "HIV/AIDS", "Population", "thinness 5-9 years", "Income composition of resources", "Schooling")])
# yy[1:1649]
head(xx)

yy <- as.numeric(yy$`Life expectancy`)
class(yy)

fit.lasso <- glmnet(xx, yy, alpha=1, standardize=TRUE)
plot(fit.lasso, label=TRUE, xvar="lambda")
title(main = 'LASSO Regression for Life Expectancy', line=3)
cbind(1:17,colnames(xx))

set.seed(1)
cv.lasso <- cv.glmnet(xx, yy, alpha=1, standardize=TRUE, nfolds=10)
plot(cv.lasso)
cv.lasso$lambda.min; log(cv.lasso$lambda.min)

lasso.coef <- coef(cv.lasso, s=cv.lasso$lambda.min)
lasso.coef
as.vector(lasso.coef)
```

No variables were shrunk to 0 using LASSO regression so it will not be considered in model selection.

## Final Model

```{r split-testing, echo = FALSE}
set.seed(1)

# Create initial split
lifeexp_split <- initial_split(lifeexp_lm_developing) # prop = 3/4 by default

# Save training data
lifeexp_train <- training(lifeexp_split)
dim(lifeexp_train)

# Save testing data
lifeexp_test  <- testing(lifeexp_split)
dim(lifeexp_test)

# Specify model
fit_train <- lm(`Life expectancy` ~  Country + Year + `Adult Mortality`
                + Alcohol + `Hepatitis B` + `under-five deaths` + `HIV/AIDS`
                + `thinness 5-9 years`+ Schooling,
                data = lifeexp_train)

fit_train
```

```{r training, echo = FALSE}
# Make predictions for training data
lifeexp_train_pred <- predict(fit_train, lifeexp_train) |>
  bind_cols(lifeexp_train |> dplyr::select(`Life expectancy`))

lifeexp_train_pred

# Calculate R-squared
rsquared <- 1 - (sum((lifeexp_train_pred - lifeexp_train$`Life expectancy`)^2) / sum((lifeexp_train$`Life expectancy` - mean(lifeexp_train$`Life expectancy`))^2))

rsquared # 0.9684521

# Calculate RMSE
class(lifeexp_train_pred)
class(lifeexp_train$'Life expectancy')

lifeexp_train_pred_numeric <- as.numeric(lifeexp_train_pred$'Life expectancy')
rmse_value <- sqrt(mean((lifeexp_train_pred_numeric - lifeexp_train$'Life expectancy')^2))

rmse_value # 0
```

```{r testing, echo = FALSE}
lifeexp_test_pred <- predict(fit_train, lifeexp_test) |>
  bind_cols(lifeexp_test |> dplyr::select(`Life expectancy`))

# Calculate R-squared
rsquared_test <- 1 - (sum((lifeexp_test_pred - lifeexp_test$`Life expectancy`)^2) / sum((lifeexp_test$`Life expectancy` - mean(lifeexp_test$`Life expectancy`))^2))

rsquared_test # 0.9626223

# Calculate RMSE
class(lifeexp_train_pred)
class(lifeexp_train$'Life expectancy')

lifeexp_train_pred_numeric <- as.numeric(lifeexp_train_pred$'Life expectancy')
rmse_value <- sqrt(mean((lifeexp_train_pred_numeric - lifeexp_train$'Life expectancy')^2))

rmse_value # 0
```

All to validate that the chosen factors are strong predictors of life expectancy. R-squared and RMSE demonstrate that this is a strong model.

